---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cyDefine 

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)

`cyDefine` robustly identifies canonical cell types as well as unassigned (and potentially novel) cells in various single-cell cytometry datasets.

## Article

The manuscript for `cyDefine` is in preparation.

Please cite with `citation("cyDefine")`.


## Install from GitHub

``` {r, eval = FALSE}
# To ensure Rstudio looks up BioConductor packages run:
setRepositories(ind = c(1:6, 8))
# Install cyCombine dependency
remotes::install_github("biosurf/cyCombine")

# Install cyDefine
remotes::install_github("claradrachmann/cyDefine", build_vignettes = TRUE)
```

## Documentation

To view the documentation for using `cyDefine`, start R and enter:

```{r vignettes, eval=FALSE}
browseVignettes("cyDefine")
```

## Usage

Please see the vignettes for a detailed description of usage.

Here is a quick run-through of the main functionalities.

```{r exclude}
library(cyDefine)
# Run initial projection to exclude redundant cell types (Optional)
reference <- excl_redundant_populations(
      reference = example_reference,
      query = example_query,
      markers = example_markers,
      min_cells = 50,
      min_pct = 0.05,
      num.threads = 4,
      mtry = floor(length(example_markers)/3),
      seed = 332
    )
```


```{r cyDefine, message=FALSE}

# Run cyDefine
classified_query <- cyDefine(
  reference = reference, 
  query = example_query, 
  markers = example_markers, 
  num.threads = 4,
  mtry = floor(length(example_markers)/3),
  num.trees = 500,
  adapt_reference = TRUE, 
  using_pbmc = FALSE, 
  batch_correct = TRUE, 
  xdim = 6, ydim = 6,
  norm_method = "scale",
  identify_unassigned = TRUE, 
  train_on_unassigned = FALSE,
  seed = 332,
  verbose = TRUE
  )
reference <- classified_query$reference
classified_query <- classified_query$query


```

## Visualizations

cyDefine provides functionality for visualizing your results by UMAP, cell type abundance, and marker expression.
Additionally, you can create a chart of the merged cell populations during reference adaptation.

```{r, fig.height=5, fig.width=12}
# Diagram of reference adaptation is only relevant if cell populations are merged (if "celltype_original" is created)
if ("celltype_original" %in% colnames(reference)) 
  plot_diagram(reference, fontcolor_nodes = c("unassigned" = "white"))
```

```{r, message=FALSE, warning=FALSE, fig.height=5, fig.width=12}

# Define a color per cell type + black for unassigned
celltype_colors <- get_distinct_colors(unique(reference$celltype), 
                                       add_unassigned = TRUE)

# UMAP of reference and query
plot_umap(reference,
          classified_query,
          example_markers,
          sample_n = 5000,
          colors = celltype_colors)
```


```{r, message=FALSE, warning=FALSE, fig.height = 7, fig.width = 13}
# Barplot of cell type abundances in query
plot_abundance(predicted_populations = classified_query$predicted_celltype,
               colors = celltype_colors)
```


```{r, message=FALSE, warning=FALSE, fig.height = 10, fig.width = 10}
# Heatmap of marker expressions per cell type in query
plot_heatmap(classified_query, 
             population_col = "predicted_celltype",
             markers_to_plot = example_markers)

```

## Report issues

If you have any issues or questions regarding the use of `cyDefine`, please do not hesitate to raise an issue on GitHub. In this way, others may also benefit from the answers and discussions.
