---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# cyDefine 

`cyDefine` robustly identifies canonical cell types as well as unassigned (and potentially novel) cells in various single-cell cytometry datasets.

## Article

A preprint of `cyDefine` is available [here](https://www.youtube.com/watch?v=dQw4w9WgXcQ&t=43s).

Please cite with `citation("cyDefine")`

## Install from GitHub

``` {r, eval = FALSE}
# To ensure Rstudio looks up BioConductor packages run:
setRepositories(ind = c(1:6, 8))
# Install cyCombine dependency
remotes::install_github("biosurf/cyCombine")

# Install cyDefine
remotes::install_github("claradrachmann/cyDefine", build_vignettes = TRUE)
```

## Documentation

To view documentation for the version of this package installed in your system, start R and enter:

```{r vignettes, eval=FALSE}
browseVignettes("cyDefine")
```


## Usage

Please see the vignettes for a detailed description of usage.

Here is a quick run-through of the functions:
