---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# cyDefine Benchmark


This repository contains the code to rerun the cyDefine benchmark.
The benchmark compares the phenotype transfer tools `cyDefine`, `CyAnno`, `Spectre`, and `CyTOF Linear Classifier` on the datasets `Levine13`, `Levine32`, `Samusik`, and `POISED`.

| Dataset | Source |
|---------|--------|
| Levine13 | HDCytoData |
| Levine32 | HDCytoData |
| Samusik | HDCytoData |
| POISED | FR-FCM-Z2V9 |

Download the POISED dataset manually. The remaining datasets will be downloaded automatically.


## Recreate figures

All results have been compiled in this repository. Figures can therefore be recreated with:

```{sh figures}
Rscript R/Figures.R
```

To rerun the entire benchmark, read on.

## Dependencies

The `CyTOF Linear Classifier` functions have been extracted and added in `R/method_CLC.R` for a simpler implementation.

### Benchmark dependecies

Besides the R packages below, CyAnno runs in a custom-built mamba/conda environment, so mamba or conda is required as well.

```{r depend}
install.packages(c("dplyr", "ggplot2", "readr", "caret", "argparse", "MASS", "patchwork", "ggpubr"))
BiocManager::install(c("HDCytoData"))
```

```{sh code_depend}
git clone https://github.com/claradrachmann/cyDefine.git
cd cyDefine
git checkout benchmark
```

### cyDefine dependecies

Install cyDefine and its dependencies

```{r cyDefine_depend}
# To ensure Rstudio looks up BioConductor packages run:
setRepositories(ind = 1:2)
# If you are correcting cytometry data, install the following Bioconductor packages:
BiocManager::install(c("flowCore", "Biobase", "sva"))
# Install cyCombine dependency
remotes::install_github("biosurf/cyCombine", ref = "dev")

# Install cyDefine
remotes::install_github("claradrachmann/cyDefine")
```

### Spectre dependecies

Install Spectre

```{r Spectre_depend}
remotes::install_github(repo = "immunedynamics/spectre")
```

### CyAnno dependecies

Clone CyAnno

```{sh CyAnno_depend}
git clone -o ../CyAnno https://github.com/abbioinfo/CyAnno.git
# Add benchmark-specific updates to the CyAnno script
cp CyAnno_changes/* ../CyAnno/

# Build environment
mamba env create -f CyAnno_changes/environment.yml
mamba activate cyanno
```

## Run the benchmark

The benchmark is orchestrated by a simple shell script.
See its content to run each module individually.

```{sh run_all}
bash src/00_run_all.sh
```


